\chapter{Sleep event detection}\label{chap:sleep-event-detection}
\begin{flushright}{\slshape 
        What, Morty, you want me to show you my math?} \\ \medskip
        --- Rick Sanchez\\Rick and Morty, season 1, episode 6
\end{flushright}
\vspace{3cm}

This chapter presents the methods developed for detection of sleep events.
The \ac{MSED} algorithm for arousal and limb movement detection, originally published in~\cite{Olesen2019}, is presented first in~\cref{sec:paperiv} and followed by the updated version in~\cref{sec:papervi}. 
A method for improving single-\ac{EEG} arousal detection is also included in~\cref{sec:paperv}. 
The chapter will conclude with a summary and discussion of the main findings of the individual research items in~\cref{sec:eventdetection-summary}.

Parts of this chapter have been modified from the following original publications:
% \paragraph{\Cref{sec:paperiv}} has been modified from\newline \fullcite{Olesen2019}\footnote{\textcopyright 2019 IEEE}
% \paragraph{\Cref{sec:paperv}} has been modified from\newline \fullcite{Olesen2020b} (\textit{submitted})
\begin{itemize}
    \item \printpublication{Olesen2019}\footnote{\textcopyright 2019 IEEE}
    \item \printpublication{Olesen2020MSED}
    \item \printpublication{Olesen2020DeepDetection}
\end{itemize}


\section{Research background}

As described in~\cref{chap:clinical-background}, a correct diagnosis of sleep disorders is predicated on precise scoring of sleep stages as well as accurate scoring of discrete sleep events. 
However, the current gold standard of manual analysis by experienced technicians is inherently biased and inconsistent due to low inter-rater reliability on the scoring of sleep stages~\cite{Norman2000,Rosenberg2013,Younes2016}, arousals~\cite{Bonnet2007}, and respiratory events~\cite{Rosenberg2014}, as described in~\cref{sec:challenges-scoring-sleep-studies}. 
As manual analysis of \acsp{PSG} is also time-consuming and prone to scorer fatigue, there is a need for efficient systems that provide deterministic and reliable scorings of sleep studies.

Although classification of sleep stages in large cohorts has been explored with good results~\cite{Olesen2018,Stephansen2018,Chambon2018c, Biswal2018, Phan2018}, reliable and consistent detection and classification of discrete \ac{PSG} events in large cohorts remains largely unexplored. 
Two studies recently proposed methods for automatic detection of arousals~\cite{Brink-Kjaer2020}, and leg movements~\cite{Carvelli2020}, and both tested their algorithms on a subset of data from two cohorts scored by multiple technicians. 
Both studies found that their algorithms could score as well as, or in some cases outperform, human scorers. 
However, both methods predicted events at discrete intervals, which might introduce biases in the decision making of when to merge and split certain predictions.

Recent studies on certain micro-events in sleep have indicated that deep learning methods reliably detect and annotate sleep spindles and K-complexes with start time and duration~\cite{Chambon2018b,Chambon2019}. 
These studies proposed a single-shot event detection algorithm, that parallels the YOLO and SSD algorithms used for object detection in \twod images~\cite{Redmon2016a,Redmon2016b,Liu2016}, but were limited in scope by detecting events only at the \ac{EEG} level, and did not explicitly take advantage of the temporal connection of the detected events. 
Additionally, experiments were carried out on a small-scale database~\cite{Chambon2018b}. 

Designing reliable and robust systems for automated sleep analysis based on machine learning algorithms often requires multiple heterogeneous data sources of sufficient size.
However, due to differences in clinical practice, few datasets in sleep science have standardized recording setups despite guidelines from the \ac{AASM}.
This creates a \textit{channel mismatch problem}, in which the overlap between our source and target domains is small, and the domains are possibly disjointed. 
Deep transfer learning\graffito{Using a pre-trained deep neural network on a separate domain.} has recently been investigated to solve the channel mismatch problem when training and testing sleep stage classification models~\cite{Phan2019, Phan2019c}.
By using a fine-tuning strategy the authors significantly improved the performance of sleep stage scoring models when trained on various combinations of \ac{EEG} and \ac{EOG} channels.

\subsection{Research motivation and objectives}

Motivated by these unresolved issues in sleep scoring, we were interested in the following research questions specifically related to research hypothesis~\ref{hypothesis:sleep-events}\graffito{\ref{hypothesis:sleep-events}: \hypothesis\xspace sleep events.}:
\newcommand{\questionSleepEventsDetection}{can sleep events be detected precisely and reliably using novel machine learning algorithms}
\newcommand{\questionSleepEventsJointly}{can the detection of one event class modulate the detection of an event from another class}
\newcommand{\questionSleepEventsTransfer}{how can we overcome the channel mismatch problem for sleep event detection}
\begin{enumerate}[label={\footnotesize\bfseries\scshape RQ~2.\arabic*}, ref={\bfseries\scshape RQ~2.\arabic*}]
    \item \questionSleepEventsDetection\label{question:sleep-events-detection}?
    \item \questionSleepEventsJointly\label{question:sleep-events-jointly}?
    \item \questionSleepEventsTransfer\label{question:sleep-events-transfer}?
\end{enumerate}
In this case, \textit{detection} covers both \textit{localization}\graffito{Localization places an unclassified sleep event in the time domain, while classification determines the class, or type, of the sleep event.} and \textit{classification} of sleep events.


% \begin{displayquote}
%     based on a single overnight \ac{PSG} recording, can we diagnose narcolepsy with the same level of performance as the current clinical gold standard?
% \end{displayquote}

Derived from the research hypothesis and associated questions, the following research objectives were formulated:

\begin{enumerate}[label=(\roman*)]
    \item a single model should detect multiple sleep events independently;
    \item the events should be annotated with a start and duration directly to avoid unnecessary postprocessing of predictions.
\end{enumerate}

The following sections describe the steps taken to complete the posed design objectives and answer the research questions.

% Research papers
\clearpage\input{chapters/research/event-detection/paper-iv.tex}
\clearpage\input{chapters/research/event-detection/paper-vi.tex}
\clearpage\input{chapters/research/event-detection/paper-v.tex}
\clearpage\input{chapters/research/event-detection/summary}